---
title: "On effect of transmission type on mpg"
author: "Jan Herman"
date: "November 22, 2015"
output: pdf_document
geometry: margin=1.8cm
---

\vspace{-1cm}
## Executive summary
\vspace{-.2cm}
The goal of this short analysis is to determine wheter the transmission
type has the effect the car fuel efficiency. If so, we aim to quantify
this effect by means of mpg difference.

We showed that the effect of transmission type on fuel efficiency
is not statistically significant based on the given data. However
the data suggests that cars with manual transmission might have
about 2 mpg better fuel efficiency compared to cars with automatic one.
We strongly recommend to collect more data in order to accept 
this hypothesis.

\vspace{-.4cm}
## Explaining the dataset
\vspace{-.2cm}
In this analysis we work with the `mtcars` dataset shipped with
the R language. It contains 11 variables describing a total of 32
different cars. The following table summarizes the meaning of
these variables:

\vspace{-.2cm}
| Variable | Description                     | Variable | Description                                   |
| -------- | --------------------------------| -------- | --------------------------------------------- |
| `mpg`    | fuel economy [miles per gallon] | `qsec`   | 1/4 mile time [sec]                           |
| `cyl`    | number of cylinders             | `vs`     | engine type (0 = vee, 1 = straight)           |
| `disp`   | engine displacement [in$^2$]    | `am`     | transmission type (0 = automatic, 1 = manual) |
| `hp`     | engine gross horsepower [hp]    | `gear`   | number of forward gears                       |
| `drat`   | rear axle ratio                 | `carb`   | number of carburetors                         |
| `wt`     | car weight [lb/1000]            |

\vspace{-.6cm}
## Goal and applied method
\vspace{-.2cm}
Our goal is to quantify the effect of the transmission type on the
fuel economy. We will try to find the most appropriate multivariable
linear model describing the relationship between `mpg` variable and
some of the other variables including `am`. The coefficient of
the `am` variable in this model will then mean the effect of
the transmission type on the fuel economy measured in mpg.

In order to make an analysis as precise as possible, we have to
preproccess the dataset a little -- we will convert variables that
have only a few integer values into factors. This involves variables
`cyl`, `gear`, `carb`, `am` and `vs`. For the last two is this
transformation unnecessary in means of the linear reggression 
(these variables have only two different values -- 0 and 1, thus
turning them into factors does not change the reggression model
at all).

```{r echo=FALSE, message=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
data("mtcars")
mtcars_factors <- mutate(mtcars, 
                         cyl = factor(cyl),
                         gear = factor(gear),
                         carb = factor(carb),
                         am = factor(am, 
                                     labels = c("automatic",
                                                "manual")),
                         vs = factor(vs, 
                                     labels = c("vee", "straight")))
```

\vspace{-.6cm}
## Choosing the model
\vspace{-.1cm}
This is the most tricky part of the entire analysis. We want to
have our model as simple as possible and at the same time it
should estimate the `mpg` variable well enough. Especially we
want to remove a majority of effect of "skewness" of the data.
In order to achive this we at first analyze the covariances of
all pairs of the variables. The following table present those.

```{r echo=FALSE}
cor_matrix <- cor(mtcars)
kable(round(cor_matrix, 3))
```

We will build a sequence of nested linear models having `mpg`
variable as outcome. The first model will be just the constant
model, the second one the simple linear regression on `am`.
Now we order the remaining variables in descending order with
respect to the absolute value of covariance with the outcome
varible `mpg` and add them successively to the model. The residual
plots of all of the models are presented in the Appendix [since
all the models have different predictors we decided to plot
on x-axis the row number in `mtcars` dataset corresponding
to the residual].

```{r echo=FALSE}
ordering_permutation <- order(abs(cor_matrix[1, ]),
                              decreasing = TRUE)
var_names <- colnames(mtcars)[ordering_permutation]
var_names <- var_names[-which(var_names %in% c("mpg", "am"))]
var_names <- c("am", var_names)
fits <- list(lm(mpg ~ 1, data = mtcars_factors))
for (i in 1:length(var_names)){
    formula_string <- paste("mpg ~ ",
                            paste(var_names[1:i],
                                  collapse = " + "),
                            sep = "")
    fit <- lm(eval(parse(text = formula_string)), 
              data = mtcars_factors)
    fits <- c(fits, list(fit))
}
```

We end with 11 different models -- the first one is the constant
model, then in each step we add one variable to the predictors
in the following order: ``r var_names``.

Now we make an analysis of variance on these nested models in order
to determine predictors that we want to keep in the model and those
we want to exclude.

```{r echo=FALSE}
models_anova <- do.call(anova, fits)
kable(cbind(`Added predictor` = c(NA, var_names),
            round(models_anova, 3)))
reduced_predictors <- var_names[(models_anova$F > 1)[-1]]
formula_string <- paste("mpg ~ ",
                        paste(reduced_predictors, collapse = " + "),
                        sep = "")
```

If we keep only predictors with $F$-statistic greater than one 
(there is a large gap) and exclude the others, we get the 
model described by the formula ``r formula_string``:

```{r echo=FALSE}
reduced_model <- lm(eval(parse(text = formula_string)),
                    data = mtcars_factors)
kable(summary(reduced_model)$coefficients)
formula_string <- paste("mpg ~ ",
                        paste(reduced_predictors
                              [-which(reduced_predictors == "cyl")],
                              collapse = " + "),
                        sep = "")
```

Finally we exclude the `cyl` predictor since its effect is not
monotonous and the $p$-value for the 8 cylinder is too high.
The final model thus has the formula ``r formula_string`` and
its summary is as follows:

```{r echo=FALSE}
final_model <- lm(eval(parse(text = formula_string)),
                    data = mtcars_factors)
kable(summary(final_model)$coefficients)
```

\vspace{-.6cm}
## Conclusion
From the table above we see that the $p$-value associated with
the `am` regressor is greater than 0.05, and so we
failed to reject the null hypotheses that the type of trasmission
has no impact on car fuel efficiency at 95% confidence level.
However the data suggests that the hypothesis might be valid,
so we higly recommend to collect more data to confirm
it definitely.

The second proposed question about the quantification of the
effect of the transmission type is thus meaningless. We hope
that the following analysis on larger data will prove our
hypotheses that manual transmissions are better for fuel
efficiency by approximately 2 mpg.

\pagebreak

## Apendix

At first present one figure -- the matrix of one variable
linear regressions together with covariances of each pair
of variables. In the main part it was replaced by covariance
table as it covers less space.

```{r echo=FALSE, message=FALSE, cache=TRUE, fig.width=7.4, fig.height=7}
library(GGally)
ggpairs(mtcars, 
        upper = list(params = list(size = 3)),
        lower = list(continuous = "smooth"))
```

\pagebreak
Next come the residual plots of all the nested models:

```{r echo=FALSE, fig.width=7.4, fig.height=9.7}
library(grid)
grid.newpage()
pushViewport(viewport(
    layout = grid.layout(nrow = (length(fits) + 1) %/% 2,
                         ncol = 2)))
for (i in 1:length(fits)){
    g <- ggplot() +
        aes(x = 1:nrow(mtcars), y = fits[[i]]$residuals) +
        geom_point() +
        labs(x = "Car number", y = "mpg residuals") +
        ggtitle(deparse(formula(fits[[i]]))) +
        theme(plot.title = element_text(size = rel(.75)),
              plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"),
              axis.title.x = element_text(size = rel(.75))) 
    print(g, 
          vp = viewport(layout.pos.row = (i + 1) %/% 2,
                        layout.pos.col = (i - 1) %% 2 + 1)
    )
}
    
```

\pagebreak
And finally the residual plots for the reduced and final model:

```{r echo=FALSE, fig.width=7.4}
ggplot() +
    aes(x = 1:nrow(mtcars), y = reduced_model$residuals) +
    geom_point() +
    labs(x = "Car number", y = "mpg residuals") +
    ggtitle(deparse(formula(reduced_model)))
    
```

```{r echo=FALSE, fig.width=7.4}
ggplot() +
    aes(x = 1:nrow(mtcars), y = final_model$residuals) +
    geom_point() +
    labs(x = "Car number", y = "mpg residuals") +
    ggtitle(deparse(formula(final_model)))
    
```

## Notice
The knitr source code of this analysis including dirty tricks
to fit the main part on 2 pages is available at
[github](https://github.com/hermitko/RegMods_CourseProject).

```{r echo=FALSE}
# Dirty tricks not to outreach 2 pages:
# 1) page margins set in knitr yaml header
# 2) 1st level headers unused - 2nd level instead
# 3) some negative vspaces added
```
